---
title: "Global Mortality"
author: "Deepal DSilva"
date: "April 16, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the required libraries
```{r}
library(tidyverse)
library(readxl)
```
Read in the data
```{r}
mortality <- read_excel("global_mortality.xlsx")
```

Summarize the data
```{r}
str(mortality)
summary(mortality)
```

Get rid of all the % in the column names
```{r}
names(mortality) <- names(mortality) %>%
  gsub("\\s\\(%\\)","", .)

mortality.long <- mortality %>% 
  gather(cause, percent, 4:35) %>%
  mutate(cause = as.factor(cause), country = as.factor(country), country_code = as.factor(country_code))

```


Leading Causes of Death around the World
```{r}
mortality.long  %>% 
  filter(year %in% c(2016), country_code %in% ("OWID_WRL")) %>%
  mutate(cause = fct_reorder(cause, percent, min)) %>%
  ggplot(aes(x = cause, y = percent, fill = cause)) + 
  geom_bar(stat = "identity") +
  coord_flip() +
  guides(fill = FALSE) + 
  ggtitle("Causes of Mortality around the World (2016)") +
  scale_y_continuous(breaks = c(0,5,10,15,20,25,30), labels = c("0%", "5%", "10%", "15%", "20%", "25%", "30%")) +
  labs(y = NULL, x = NULL) +
  theme_minimal()

ggsave("Plot1.png", width = 30, height = 20, units = "cm")
 
```
 
################################

I calculated the percent difference between years 1990 and 2016 (for the World data) to find the change in percent over the years. 
Then I sorted the causes based on the increase or decrease change in percent.

```{r}

#Data Prep
mortality_world <- mortality.long  %>% 
  filter(country_code %in% ("OWID_WRL"), year %in% c(1990, 2016)) %>%
  spread(year, percent) %>%
  mutate(percentchange = `2016` - `1990`) 

mortality_world$percentchange <- round(mortality_world$percentchange, digits = 2)
mortality_world$percent_type <- ifelse(mortality_world$percentchange < 0, "decreased", "increased") 

mortality_world <- mortality_world[order(mortality_world$percentchange),]
mortality_world$cause <- factor(mortality_world$cause, levels = mortality_world$cause)
  

  #Diverging Bar Chart
  
  ggplot(aes(x = cause, y = percentchange, label = percentchange), data = mortality_world) + 
  geom_bar(stat = "identity", aes(fill = percent_type), width = .5) +
  geom_text(color = "black", size = 2.5, nudge_y = 0.2) +  
  coord_flip() +  
  scale_fill_manual(name="Percent Change", 
                    labels = c("Increased", "Decreased"), 
                    values = c("increased"="#00ba38", "decreased"="#f8766d"))  +  
  labs(y = NULL, x = NULL, title = "Causes of Mortality around the World", subtitle = "Change in Percent from 1990 to 2016") +
  theme_minimal()
  

  ggsave("Plot2.png", width = 30, height = 20, units = "cm")  
```
 We've seen a sharp increase in Cardiovascular diseases, cancers and diabetes while neonatal deaths and diarrheal disease deaths have reduced the most over the years.
 
 ####################################
 
 Just curious about how India stands, so created this chart
 
```{r}
mortality.long  %>% 
  filter(year %in% c(2016), country_code %in% ("IND")) %>%
  mutate(cause = fct_reorder(cause, percent, min)) %>%
  ggplot(aes(x = cause, y = percent, fill = cause, frame = year)) + 
  geom_bar(stat = "identity") +
  coord_flip() +
  guides(fill = FALSE) + 
  ggtitle("Causes of Mortality in India (2016)") +
  scale_y_continuous(breaks = c(0,5,10,15,20,25,30), labels = c("0%", "5%", "10%", "15%", "20%", "25%", "30%")) +
  labs(y = NULL, x = NULL) +
  theme_minimal()
```

